---
title: "Generalized Estimating Equations"
author: "Ashley Dosch, Amy Muller, Anthony Stawiery"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advance Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
editor: 
  markdown: 
    wrap: 72
---

Cumluative references and summaries: https://docs.google.com/document/d/1h3yuMl3h2s_q4uzmoouksLEmx3fMcz9ML1uDiyCilMI/edit?usp=sharing

## Introduction

  Generalized Estimating Equations (GEE) is a statistical methodology commonly used to analyze correlated data in various fields that use panel or longitudinal data, including epidemiology, social sciences, and clinical trials. GEE extends the generalized linear model framework to account for within-cluster correlations, such as repeated observations or clustered data. GEE is used to model the marginal mean response while allowing for flexible correlation structures among observations within the same group or cluster. GEE estimates population-average effects by specifying a working correlation structure such as independence or exchangeable structures. GEE also implements an estimating equation approach to iteratively update the parameter estimates.
    
  This methodology provides reliable and robust estimates, even when the correlation structure is misspecified. GEE is particularly useful for longitudinal or panel data analysis, as it allows for examination of all available data in the experiment and provides statistical inference for population-level ramifications. Its ability to handle missing data makes GEE a great choice for analyzing correlated data, particularly in biostatistics or research settings.


## Methods

**(Data Structure)**: GEE is designed for analyzing correlated data, such as longitudinal or panel data. The data are organized into clusters, where each cluster represents a group of related observations. An example is repeated measures on the same patient for medicinal results over time.

**(Model Specification)**: GEE uses a generalized linear model (GLM) framework to specify the relationship between the response variable and the regressor variables. The choice of the link function and the distribution of the response variable depends on the nature of the data.

**(Correlation Structure)**: GEE allows for flexible modeling of the correlation structure within clusters. A working correlation structure is specified, which represents the assumed relationship among the correlated observations. The correlation structures discussed in this paper are exchangeable, autoregressive, unstructured, and independence.

**(Estimating Equations)**: Instead of maximizing the likelihood function, GEE employs an estimating equation approach to estimate the regression parameters. The estimating equations are derived based on the assumed correlation structure and the marginal mean model. These equations capture the population-averaged effects of the explanatory variables.

**(Iterative Algorithm)**: GEE uses an iterative algorithm, typically based on the Newton-Raphson method, to update the parameter estimates. The process involves solving the estimating equations iteratively until convergence is reached.

**(Inference)**: GEE provides robust standard errors for the estimated regression parameters, taking into account the correlation structure. Hypothesis tests and confidence intervals can be constructed to assess the statistical significance of the variables and make inferences about the population-level effects.

**(Model Assessment)**: After fitting the GEE model, various model diagnostics can be performed to assess the model's goodness of fit, such as residual analysis, model comparison criteria (e.g., Akaike Information Criterion), and assessing the adequacy of the assumed correlation structure. This can be thought of as using the Linear Regression analysis we did in previous stats class


## Analysis and Results

### Data and Vizualisation

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}

# loading packages 
library(tidyverse)
library(knitr)
library(geepack)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
data(respiratory)

# regular GLM
m1 <- glm(outcome ~ center + treat + age + baseline, data=respiratory,                
          family=binomial())     

# GEE with independence correlation structure
gee_independence <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family=binomial(), corstr="independence")

# GEE with exchangeable correlation structure
gee_exchangeable <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family=binomial(), corstr="exchangeable")

# GEE with unstructured correlation structure
gee_unstructured <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family=binomial(), corstr="unstructured")

# GEE with Auto-Regressive Order 1 correlation structure
gee_ar1 <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family=binomial(), corstr="ar1")  


# Creating list data structure for all relevant correlation types (different GEE models)
corrStructureList = list(gee_independence, gee_exchangeable, gee_unstructured, gee_ar1)

# Loops through the list and performs "row binding {rbind}" to form a single batch
# lapply + QIC = "Quasi Information Criterion" which allows for quasi-likelihood under the independence model
do.call(rbind, lapply(corrStructureList, QIC))


lapply(corrStructureList, tidy)

# https://online.stat.psu.edu/stat504/lesson/12/12.1 (for correlation structure information)
  

```

### Statistical Modeling

### Conlusion

## References


#  McCulloch, Searle, S. R., & Searle, S. R. (Shayle R. . (2008). Generalized, linear, and mixed models. John Wiley & Sons, Ltd.
