---
title: "Generalized Estimating Equations"
author: "Ashley Dosch, Amy Muller, Anthony Stawiery"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advance Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
nocite: '@*'
editor: 
  markdown: 
    wrap: 72
---

**Literature Review**: https://docs.google.com/document/d/1h3yuMl3h2s_q4uzmoouksLEmx3fMcz9ML1uDiyCilMI/edit?usp=sharing

## Introduction

  Generalized Estimating Equations (GEE) is a statistical methodology commonly used to analyze correlated data in various fields that use panel or longitudinal data, including epidemiology, social sciences, and clinical trials @Mcculloch2008. The method was first introduced by Liang and Zeger in a paper entitled "Longitudinal Data Analysis Using Generalized Linear Models" in 1986. @Liang1986. Analysis of longitudinal data has inherent problems with independence involving multiple observations for the same individual in clinical data studies. Liang and Zeger even referred to time dependence on multiple observations of an individual as a nuisance due to correlation issues. @Liang1986. Another challenge for use of longitudinal data is the need for more complex statistical models to deal with the lack of independence in observations that could include multidimensional levels of integration, and the software is often not available.@donald2006longitudinal The use of GEE mitigates these challenges with use of longitudinal data.
  
  GEE extends the generalized linear model framework to account for within-cluster correlations, such as repeated observations or clustered data @shao2023comparison. GEE is used to model the marginal mean response while allowing for flexible correlation structures among observations within the same group or cluster. GEE estimates population-average effects by specifying a working correlation structure such as independence or exchangeable structures @shao2023comparison. GEE also implements an estimating equation approach to iteratively update the parameter estimates.
    
  This methodology provides reliable and robust estimates, even when the correlation structure is mis-specified. GEE is particularly useful for longitudinal or panel data analysis, as it allows for examination of all available data in the experiment and provides statistical inference for population-level ramifications @lin2011impact. Its ability to handle missing data makes GEE a great choice for analyzing correlated data, particularly in biostatistics or research settings.

 General mathematical formula for multivariate GEE: @PSU2023.
$$
 \log\left(\dfrac{\pi_{ij}}{1-\pi_{ij}}\right)=\beta_0+Regressor1_{ij}\beta_1+Regressor2_{ij}...\beta_2+RegressorN\beta_{N-1}
$$

  In this review of GEE, the 'geepack' package for R with a respiratory dataset included was used to showcase an example. @Hojsgaard2006 @Yan2004 @Yan2002 The respiratory dataset included in the packet is an example of longitudinal or panel data as data is from a clinical study of 111 patients receiving treatment for a respiratory illness.@koch1990categorical 

## Methods

### Data Structure 
GEE is designed for analyzing correlated data, such as longitudinal or panel data. The data are organized into clusters, where each cluster represents a group of related observations. An example is repeated measures on the same patient for medicinal results over time.

### Model Specification 
GEE uses a generalized linear model (GLM) framework to specify the relationship between the response variable and the regressor variables @shao2023comparison. The choice of the link function and the distribution of the response variable depends on the nature of the data @gayen2018generalized.

### Correlation Structure 
GEE allows for flexible modeling of the correlation structure within clusters. A working correlation structure is specified, which represents the assumed relationship among the correlated observations @seals2016analysis. The correlation structures discussed in this paper are exchangeable, autoregressive, unstructured, and independence.
 

### Estimating Equations 
Instead of maximizing the likelihood function, GEE employs an estimating equation approach to estimate the regression parameters @d2011generalized. The estimating equations are derived based on the assumed correlation structure and the marginal mean model @d2011generalized. These equations capture the population-averaged effects of the explanatory variables.

### Iterative Algorithm 
GEE uses an iterative algorithm, typically based on the Newton-Raphson method, to update the parameter estimates. The process involves solving the estimating equations iteratively until convergence is reached. {needs more sourcing / description}

### Inference 
GEE provides robust standard errors for the estimated regression parameters, taking into account the correlation structure. Hypothesis tests and confidence intervals can be constructed to assess the statistical significance of the variables and make inferences about the population-level effects @shao2023comparison.

### Model Assessment 
After fitting the GEE model, various model diagnostics can be performed to assess the model's goodness of fit, such as model comparison criteria and assessing the adequacy of the assumed correlation structure @shao2023comparison.


## Analysis and Results

### Data and Vizualisation
https://cran.r-project.org/web/packages/geepack/geepack.pdf

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(geepack)
library(lme4)   # For fitting linear mixed models
library(glmmTMB)  # For fitting GLMMs with different distributions
library(deSolve)

# Load Data
data(respiratory)

# Data Information/Visual
str(respiratory)
head(respiratory)


```
  GEE formula for the respiratory data:
$$
 \log\left(\dfrac{\pi_{ij}}{1-\pi_{ij}}\right)=\beta_0+Center_{ij}\beta_1+Treat_i\beta_2+Age_i\beta_3+Sex_i\beta_4+Baseline_i\beta_5
$$

### Statistical Modeling
```{r}
# GLM Model
glm_example <- glm(outcome ~ center + treat + age + baseline + sex, data=respiratory,                
          family = binomial())

summary(glm_example)

# GLMM Model
glmm_example <- glmmTMB(outcome ~ center + treat + age + baseline + (1 | id),
                 data = respiratory, family = binomial())

summary(glmm_example)


# Mixed Markov Example


# Differential Equation Example



# GEE with independence correlation structure
gee_independence <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family = binomial(), corstr="independence")

# GEE with exchangeable correlation structure
gee_exchangeable <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family = binomial(), corstr="exchangeable")

# GEE with unstructured correlation structure
gee_unstructured <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family = binomial(), corstr="unstructured")

# GEE with Auto-Regressive Order 1 correlation structure
gee_ar1 <- geeglm(outcome ~ center + treat + age + baseline, data=respiratory, id=id, 
                  family = binomial(), corstr="ar1")  


# Displays table of QIC between all 4 different types of GEE models with different
# correlation structure
QIC(gee_independence, gee_exchangeable, gee_unstructured, gee_ar1)
```

### Conclusion

{Let's talk about quasi-likelihood and reasons why we'd choose GEE over other models?}

## References



