---
title: "Generalized Estimating Equations"
author: "Ashley Dosch, Amy Muller, Anthony Stawiery"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advance Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
nocite: '@*'
editor: 
  markdown: 
    wrap: 72
---

## Introduction

Generalized Estimating Equations (GEE) is a statistical methodology in a
class of models that model the mean of an outcome as function of
predictor variables commonly used to analyze correlated data in various
fields that use panel or longitudinal data, including epidemiology,
social sciences, and clinical trials [@Mcculloch2008]. The method was
first introduced by Liang and Zeger in a paper in 1986.
[@Liang1986longitudinal]. The method uses a quasilikelihood approach and
yields population averaged estimates of parameters.
[@wang2014generalized]

Analysis of longitudinal data has inherent problems with independence
involving multiple observations for the same individual in clinical data
studies. Liang and Zeger even referred to time dependence on multiple
observations of an individual as a nuisance due to correlation issues.
[@Liang1986longitudinal]. Another challenge for use of longitudinal data
is the need for more complex statistical models to deal with the lack of
independence in observations that could include multidimensional levels
of integration, and the software is often not
available.[@donald2006longitudinal] The use of GEE mitigates these
challenges with use of longitudinal data.

GEE extends the generalized linear model framework to account for
within-cluster correlations, such as repeated observations or clustered
data. [@shao2023comparison] GEE is used to model the marginal mean
response while allowing for flexible correlation structures among
observations within the same group or cluster. GEE estimates
population-average effects by specifying a working correlation structure
such as independence or exchangeable structures.[@shao2023comparison]
GEE also implements an estimating equation approach to iteratively
update the parameter estimates.

This methodology provides reliable and robust estimates, even when the
correlation structure is mis-specified.[@wang2014generalized] GEE is
particularly useful for longitudinal or panel data analysis, as it
allows for examination of all available data in the experiment and
provides statistical inference for population-level ramifications.
[@lin2011impact] Its ability to handle missing data makes GEE a great
choice for analyzing correlated data, particularly in biostatistics or
research settings without the use of intense statistical mathematics.

## Methodology

### Estimating Equations

Instead of maximizing the likelihood function, GEE employs a
quasilikelihood estimating equation approach to estimate the regression
parameters.[@d2011generalized] The estimating equations are derived
based on the assumed correlation structure and the marginal mean
model,not a specified distriution. [@d2011generalized] GEE models can be
for continuous, binary or count outcomes.

### Iterative Algorithm

GEE uses an iterative algorithm after using initial estimates for regression coefficients, $\beta_i$, to update the parameter estimates. Software packages, such as 'geepack' package in R handle the iteration process. The iterative process works until there is convergence to stable estimates. [@Hojsgaard2006] [@Liang1986longitudinal] 

### Data Structure

GEE is designed for analyzing correlated data, such as longitudinal or
panel data in which the data is organized into clusters, where each
cluster represents a group of related observations.[@Mcculloch2008] The assumption of independent observations may not be accurate. A GEE analysis of a dataset with longitudinal data with repeated observations per individual will be shown using the respiratory dataset available through the
'geepack' package in R.

### Model Specification

GEE is an extension of generalized linear models (GLM) and uses that
framework to specify the relationship between the response variable and
the regressor variables. [@shao2023comparison] The choice of the link
function and the distribution of the response variable depends on the
nature of the data.[@gayen2018generalized] Models can be created for
continuous, binary or count outcomes. The table below shows the
extension of GEE from the LM, GLM,and GLMM and differences in assumptions.

![Figure 1: Model Comparision and Summary [@Mcculloch2008]](Model_Comparison_Table.jpg)

According the Liang and Zeger, the method of GEE, longitudinal data
sets, are comprised of an outcome variable, $y_{it}$, and
\$\$p\\times1\$\$ vector of covariates, $x_{it}$, observed at times
$$t=1,...,n_{it}$$ for subjects $$i=1,...,K$$ which arise often in
applied sciences. [@Liang1986longitudinal] For example, in health care
study data, individual patients have follow visits to monitor a
treatment plan and therefore have multiple observations for each
patient. GEE is used for these kind of longitudinal datasets due to the
use of correlation structures in its modeling design to account for the
lack of independence in the repeated observations.

In Wang's review of GEE method, a marginal model specifying a
relationship between $\mu_i$ and $X_{ij}$ is given by the equation
$g(\mu_{ij})=X'_{ij} \beta$, where $g$ is the link function and $\beta$
is a vector of regression coefficients and then shows the mathematical
calculations to get to the equation

$U(\beta) = \sum_{i=1}^K {D_i}^{\prime} {V_i}^{-1}(Y_i-\mu_i)=0$

that when solved will give estimates of $\beta_i$ used in the generated
models. In this equation $D_i=\partial\mu_i/\partial\beta^{\prime}$ and
${V_i}^{-1}$ is the inverse of the variance-covariance matrix for $Y_i$,
$V_i = \phi{A_i}^{1/2}R_i(\alpha){A_i}^{1/2}$ , $A_i$ is a diagonal
matrix with elements that are known variance functions of $\mu_{ij}$
times a scaling parameter $\phi$, and $R_i(\alpha)$ is the correlation
structure. [@wang2014generalized]

GEE employ a logistic regression model created using the 'geepack'
package available in R, specifying a correlation structure (see the
structure types in the paragraph below), family type of binomial for
discrete outcomes and would yield model of

General mathematical formula for multivariate GEE using the logit link
function for binary outcomes: 
$$
 \text{logit}[P(y_\text{ij}=1)]=log\left(\dfrac{\pi_{ij}}{1-\pi_{ij}}\right)=\beta_0+Regressor1_{ij}\beta_1+Regressor2_{ij}\beta_2...+RegressorN\beta_{N-1}
$$

that predicts log odds ratios for outcomes for each predictor.
[@PSU2023]

The 'geepack' package in R uses the link and variance functions in the
geeglm() function of identity for gaussian, logit and
$\mu(1-\mu),\mu\in(0,1)$ for binomal, log and$\mu,\mu>0$ for poisson,
and log and $\mu^2,\mu>0$ for gamma families. [@Hojsgaard2006]

### Correlation Structures

GEE allows for flexible modeling of the correlation structure within
clusters. A working correlation structure is specified, which represents
the assumed relationship among the correlated observations
[@seals2016analysis]. The correlation structures discussed in this
paperare exchangeable, autoregressive (AR1), unstructured, and
independent.

![Figure 2: Correlation Structures Comparison Table
[@wang2014generalized]](corr-table.jpg)

The independent correlation structure is used when the assumption
between multiple repeated observations of a single individual is
independent. Use of the exchange correlation structure is used when
there is equality between times of observations. Autoagressive (AR1)
correlation structure is used when there are more than one set of equal
time differences between repeated observations. And lastly, unstructured
correlation structure is the most general and takes into account all the
possible time intervals between repeated observations. Of course there
are more choices for correlation structures, but these (4) structures
are the structures used in the 'geepack' R package used in the
subsequent examples.

If the correlation structure chosen for GEE is not correct, the
predictor variables will still be estimated consistently, since the GEE
method relies on the first moment, however the standard errors will be
incorrect. This considered very attractive property of GEE
models.[@donald2006longitudinal] A possible remedy for choosing a
correlation structure incorrectly is to use the 'sandwich' package in R
that uses the Huber-White sandwich estimator to create an an
approximation for the correlation.[@homish2010analyzing]
[@zeileis2019package] By default, the 'geepack' package in R uses a
sandwich estimator if the number of clusters is larger than 30, but has
a built in options to use another variance estimator called the
jackknife estimator, fully iterated or approximate, for cases where the
number of clusters is less than 30 (as recommended by
[@paik1988repeated]).[@Hojsgaard2006]

Correlation structure selection is determined through various methods
and is not limited to the structures shown in the correlation table
above. One valid method used to determine model selection has been
proposed by Shults and Chaganty (@wang2012penalized) as to determine the
structure based on the generalized error sum of squares given by the
following equation: $$
  ESS(\alpha,\beta) = \sum_{i=1}^{K}{(Y_i - \mu _i)}'{V_{i}}^{-1}(Y_i - \mu _i)
$$. In addition to the method by Schuts and Chaganty, Wang discusses
several other methods for correlation structure selection and cites many
other methods available. [@wang2014generalized]

## Analysis and Results

### Data and Vizualisation

In this review of GEE, the 'geepack' package for R with a respiratory
dataset included was used to showcase an example. [@Hojsgaard2006]
[@Yan2004] [@Yan2002] The respiratory dataset included in the packet is
an example of longitudinal or panel data as data is from a clinical
study of 111 patients receiving treatment for a respiratory illness.
[@koch1990categorical]

| Variable | Description                    |
|----------|--------------------------------|
| center   | a numeric vector               |
| id       | a numeric vector               |
| treat    | treatment or placebo           |
| sex      | male or female                 |
| age      | in years at baseline           |
| baseline | respiratory status at baseline |
| visit    | id of visit                    |
| outcome  | respiratory status at visit    |

: {tbl-colwidths="\[20,35\]"}

```{r, warning=FALSE, echo=T, message=FALSE}
######## loading packages ########

library(tidyverse)
library(knitr)
library(geepack)
library(lme4)   # For fitting linear mixed models
library(glmmTMB)  # For fitting GLMMs with different distributions
library(ggeffects)
library(dplyr)
library(ggplot2)
library(furniture)

######## Load Data ########

data(respiratory)

displayData <- respiratory[respiratory$id < 4,]
displayData <- displayData[displayData$center == 2,]

######## Plot sample of data for visualization ########

ggplot(displayData, aes(x=visit, y=treat, color=factor(outcome))) +
  geom_point(size=5) +
  scale_x_continuous(breaks=c(1,2,3,4)) +
  scale_color_discrete("Respiratory Health",labels=c("Poor","Good")) +
  facet_grid(.~id) +
  labs(x = "Visit Number/Patient", y = "Treatent, 'A' = Treatment, 'P' = Placebo",title = "Respiratory Treatment Outcomes Over (4) Patient Visits")+
  theme_bw()
```

The above graphic is displaying a small sample from the 'respiratory'
data. The given data set is a study with longitudinal data of repeated
measures with a binary outcome. There are four measures, or visits,
performed on each individual. Each box with 1 through 4 on the x-axis
represents a unique patient and their respective visits. The y-axis
displays the treatment the patient is taking where P represents a
placebo and A represents the actual treatment.

The points on the plot indicate the outcome of treatment and represent a
patient's respiratory status per visit where 1 indicates a "good"
respiratory status and 0 indicates a "poor" respiratory status. Note
that this graphic is a truncated form of the complete dataset and
excludes both a treatment center and 52 other patients.

```{r, warning=FALSE, echo=T, message=FALSE}
###### Data transformation for "wide" data [record per visit] #######
resp_transformed <- respiratory %>% 
  tidyr::spread(key = visit,
                value = outcome,
                sep = "_") %>%
  dplyr::arrange(id) %>% 
  dplyr::select(id, center, 
                sex, age, treat, 
                baseline, starts_with("visit"))
```

'resp_transformed' is a data frame that has been transformed based on
the original data set. This transformation manipulates the data such
that there is one line per participant in the data. This method will
allow for analyzing the metrics between repeated observations.

```{r, warning=FALSE, echo=T, message=FALSE}
###### Table of data breakdown - respiratory outcome over time #######
resp_transformed %>% 
  dplyr::group_by(treat) %>% 
  furniture::table1("Visit One" = visit_1, 
                    "Visit Two" = visit_2, 
                    "Visit Three" = visit_3, 
                    "Visit Four" = visit_4, 
                    caption = "Respiratory Outcome Over Time",
                    output = "markdown",
                    na.rm = FALSE,
                    total = TRUE,
                    test = TRUE)
```

Table "Respiratory Outcome Over Time" shows a breakdown of visits and
the corresponding statistics. N is the total number of patients in the
study, "A" represents the treatment, and "P" represents the placebo.

```{r, warning=FALSE, echo=T, message=FALSE}
###### Graph of Number of Visits with Poor Outcome #######
resp_transformed %>% 
  dplyr::mutate(n_good = furniture::rowsums(visit_1 == "0", 
                                            visit_2 == "0",
                                            visit_3 == "0",
                                            visit_4 == "0")) %>% 
  ggplot(aes(x = age,
             y = n_good)) +
  geom_count() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age in Years",
       y = "Number of Visits out of Four, with 'Poor' Respiration")
```

This visualization is focused on the age parameter in terms of outcome
per visit. Based on age for the x-axis, it's displaying whether
respective patients had a 'poor' outcome (or poor respiration status) as
a number out of the four total visits. The larger circles indicate a
higher number of poor outcomes in total, while the smaller circles
indicate a smaller number of poor outcomes in total. This visual does
not display patients who had all 'good' outcomes.

```{r, warning=FALSE, echo=T, message=FALSE}
###### Correlation between repeated observations #######
resp_transformed %>% 
  dplyr::select(starts_with("visit")) %>% 
  dplyr::mutate_all(function(x) x == "1") %>% 
  cor() %>% 
  corrplot::corrplot()
```

The preceding graphic shows the correlation matrix between repeated
observations. It's noted that there is a medium to high correlation
between the observations. This is showing the necessity for a
statistical model that can account for these levels of correlation
between the observations.

### Statistical Modeling


#### GEE Model(s)

GEE models focus on the mean of the correlated observations that are within groups and don't focus on the distribution of the full data of the population [@Hojsgaard2006]. GEE modeling requires parameters in the form of distribution of the response variable, the link function, the regressor variables, and the desired correlation structure. The various types of correlation structures were discussed in the previous methodology section; our example focuses on independent, exchangeable, unstructured, and auto-regressive order 1 correlation structures. The distribution selected for a model is based on the distribution of the response variable [@seals2016analysis] and may be a value such as binomial, Poisson, or Gaussian [@gayen2018generalized]. The link function parameter is contingent on the corresponding distribution where the distribution of the response variable dictates the model to be used [@agresti2015foundations].

The 'geepack' package in R was used to create four different models using the same geeglm() function and specifying different correlation structures as parameters within the function input. The binomial distribution is chosen for each model as well due to the binary response of a 'poor' or 'good' respiratory status. The canonical link function used for binary response data is 'logit' or logistic regression [@agresti2015foundations]. This link function is inherently used in our models without direct specification.


1.  **GEE with Independent Correlation Structure**

```{r}
# GEE with independent correlation structure
gee_independence <- geeglm(outcome ~ center + treat + age + baseline + sex, data=respiratory, id=id, 
                  family = binomial(), corstr="independence")
summary(gee_independence)
```

2.  **GEE with Exchangeable Correlation Structure**

```{r}
# GEE with exchangeable correlation structure
gee_exchangeable <- geeglm(outcome ~ center + treat + age + baseline + sex, data=respiratory, id=id, 
                  family = binomial(), corstr="exchangeable")
summary(gee_exchangeable)
```

3.  **GEE with Unstructured Correlation Structure**

```{r}
# GEE with unstructured correlation structure
gee_unstructured <- geeglm(outcome ~ center + treat + age + baseline + sex, data=respiratory, id=id, 
                  family = binomial(), corstr="unstructured")
summary(gee_unstructured)
```

4.  **GEE with Auto-Regressive Order 1 Correlation Structure**

```{r}
# GEE with Auto-Regressive Order 1 correlation structure
gee_ar1 <- geeglm(outcome ~ center + treat + age + baseline + sex, data=respiratory, id=id, 
                  family = binomial(), corstr="ar1")
summary(gee_ar1)
```

### Model Assessment

After fitting the GEE model, various model diagnostics can be performed to assess the model's goodness of fit, such as model comparison criteria and assessing the adequacy of the assumed correlation structure.
[@shao2023comparison]

#### Respiratory Data GEE Formula

$$
 \log\left(\dfrac{\pi_{ij}}{1-\pi_{ij}}\right)=\beta_0+Center_{ij}\beta_1+Treat_i\beta_2+Age_i\beta_3+Sex_i\beta_4+Baseline_i\beta_5
$$

#### Plot regarding random effects of GEE Models

```{r}
plot(ggemmeans(gee_independence, terms = c("center", "treat"),
               condition = c(outcome = "poor"))) + 
  ggplot2::ggtitle("GEE Effect plot")
```

-   Do we want to keep this?

#### QIC (Quasi Information Criterion) for Model Selection\*\*

```{r}
# Displays table of QIC between all 4 different types of GEE models with different
# correlation structure
QIC(gee_independence, gee_exchangeable, gee_unstructured, gee_ar1)
```

In order to evaluate which correlation structure is best to use with GEE models, QIC, quasilikelihood information criterion, must be used instead of the AIC, Akaike's information criterion, to assess since GEE is not based on a specified distribution. AIC uses the asymptotic nature of the maximum likelihood estimator for a specified distribution and is therefore not applicable for GEE. [@pan2001akaike] When comparing the different correlation structures used for GEE, the smallest QIC value is the indicator for which is the best correlation structure to use. For the respiratory dataset, the unstructured correlation structure has the lowest QIC value of 512 when compared to the other correlation structures.Therefore, use of the unstructured correlation structure should be used for the GEE model for the respiratory dataset.

### Inference

GEE provides robust standard errors for the estimated regression parameters, taking into account the correlation structure. When comparing multiple different techniques, it's important to consider the correlation between observations in order to perform accurate
statistical inference of the model's output. [@seals2016analysis] Hypothesis tests and confidence intervals can be constructed to assess the statistical significance of the variables and make inferences about the population-level effects. [@shao2023comparison]

### Conclusion

GEE is a valid choice of statistical modeling to be used in the case
that a standard regression model is not applicable due to the
between-observation relationship within data. This directly applies to
research cases and other applications that utilize clustered or
longitudinal data. If a GEE model is not used and the between-segment
correlation is ignored, then there may be an incorrect conclusion of
significant results, or type I errors. [@seals2016analysis]

GEE uses quasilikelihood estimation which is opposite of other models
that use maximum likelihood estimation or the least squares methods.
[@shao2023comparison] To summarize, GEE is a good extension of GLM for
modeling longitudinal or panel data and data with binary or continuous
outcomes. The fact that the individual observations do not have to be
independent or a distribution specified combined with the fact that the
models are based on the first moment only and use of one several
possible correlation structures make it a good choice for modeling.

There are a few other packages available in R to create GEE models not discussed in this review GEE. They are the 'gee' package [@Carey2022] and 'multgee' pagkage specifically for multinomial responses [@touloumis2014r].


##### GLM Model

```{r}
# GLM Model
glm_example <- glm(outcome ~ center + treat + age + baseline + sex, data=respiratory,                
          family = binomial())
summary(glm_example)
```

GLM, shown as the glm() function above, does not account for the between
segment correlations in longitudinal data. We are accounting for the
distribution for this model (where that is defined in the family =
binomial() parameter), but it is still not sufficient to account for the
between-segment correlation. In this case we are specifying that the
model is a logistic regression model.

Using the variables center, treatment, age, baseline, and sex, the model
is attempting to predict the response variable value of "1" for outcome.
This would indicate a 'good' respiratory status.

The p-values found in the coefficients table of the summary indicate
that treatment and baseline are good predictors for respiratory outcome.
Center and age are decent predictors as well. Sex, on the other hand, is
not a good predictor for our given responses.

##### GLMM Model

```{r}
# GLMM Model
glmm_example <- glmmTMB(outcome ~ center + treat + age + baseline + sex + (1 | id),
                 data = respiratory, family = binomial())
summary(glmm_example)
```

GLMM, or generalized linear mixed effect model, is also used here to
determine how the predictor variables affect the respiratory status
outcome. A GLMM may be used when there are multiple levels of
observations in the data and the end goal is to further investigate the
relationship between variables within a level [@shao2023comparison]. As
an extension of GLM, GLMM uses interactions between time and regressor
variables to determine the change of a response variable over time. GLMM
is an appropriate method of examining longitudinal data in the case that
both within-patient and between-patient differences are the research's
end goal [@shao2023comparison].

Using the variables center, treatment, age, baseline, and sex, the model
is attempting to determine how these variables affect the likelihood of
a 'poor' outcome versus a 'good' outcome within a patient's visit.

The p-values found in the coefficients table of the summary indicate
that treatment, center, and baseline are good predictors for respiratory
outcome. Age and sex are not good indicators in this model for
predicting respiratory outcomes. Is it important to note that there are
no listed random effects which indicates that there is not sufficient
individual variability to account for them [@shao2023comparison].

## References

## Literature Review

[Generalized Estimating Equations Literature
Review](https://docs.google.com/document/d/1h3yuMl3h2s_q4uzmoouksLEmx3fMcz9ML1uDiyCilMI/edit?usp=sharing)
